---
title: "GO-Analysis-with-HypeR"
output: html_document
---
## Geneontology analysis

HypeR strores te information of diverent data bases for several organisms:
```{r hypeR_datasets, include=T, eval = F}
library(hypeR)
# hypeR strores te information of diverent data bases for several organisms
msigdb_info()
```

From this we can load for example the REACTOME pathway information:

```{r load_REACTOME_geneset, include=T, eval = F}
# get pathway info from reactome
reactome_geneset <- msigdb_gsets("Homo sapiens", "C2", "CP:REACTOME")
```

Now we want to know how many of the genes we found to be bound, belong to which of the reactome pathways. For this, we make a list of the names of all bound genes and test for the overlap:

```{r reactome_overlap, include=T, eval = F}
# get names of bound genes
bound_gene_names <- hit_genes$gene_name

# test enrichment of genes in reactome pathways
reactome_hyp <- hypeR(bound_gene_names, reactome_geneset, test="hypergeometric", fdr=0.05)

reactome_hyp

# Plot dots plot
reactome_dot <- hyp_dots(reactome_hyp)+
  ggtitle("bla")

reactome_dot

```

**Exercise: GO termes**
Another often used data base is Geneontology. Geneontology stores three types of information: Cellular Components, Molecular Functions and biological progesses. All three are also availabe as genesets from hypeR (See msigdb_info()). Make the corrisponding dotplots for these genesets.