---
title: "Visualizing data with ggplot2"
date: last-modified
author: 
- name:  "Natascha Kleinermanns & Zarnack group"
title-block-banner: true
format: 
    html:
        theme: litera
        self-contained: true
        code-fold: false
        toc: true
        number-sections: false
        anchor-sections: false
        fig-format: png
        embed-resources: true
execute:
  include: true
  eval: true
  echo: true
  error: false
  message: false
  warning: false
editor: 
  markdown: 
    wrap: 100
---
# What is ggplot2 - The Basics

`ggplot2` is a package for creating the advanced graphics in R. You provide the data, tell ggplot2 
how to map variables to aesthetics, what graphical primitives to use, and it takes care of the 
details. The ggplot2 package offers a modular structure that allows you to generate plots layer by layer.

## Installing ggplot2

For installing packages, we can use the `install.packages()` function. After the package installation, we need to load the package into our R environment, so we can use 
all the functions inside the package.

```{r, eval=FALSE}
install.packages("ggplot2")
```
```{r}
## load the ggplot2 package
library(ggplot2)
```

## The basic structure of ggplot2

The most important function is `ggplot()` that initializes a ggplot object in which we declare the input data frame and set the plot aesthetics. The aesthetics are defined within `ggplot()` by using `aes()`. Aesthetics tell ggplot which columns contain the values on the x- and y-axis or which column should be used as a color or fill. After we have initialized a ggplot object with `ggplot()` we can add layers with the plus sign `+`. Components like points, lines, bars etc. can be added to a plot with `geom_*()` functions. There are further functions that can be used to adjust the x- and y-axis, to set labels or to set a user-defined color-scale. Here are some examples of layers:

-   geom_point() for scatter plots
-   geom_line() for line plots
-   geom_col() for bar charts
-   geom_histogram() for histograms
-   geom_density() for density plots
-   facet_wrap() for dividing a plot in multiple subplots
-   labs() for renaming and adding axis labels and titles
-   scale_color_manual() for setting a user defined color-scale

For a comprehensive list of `geom_*()` and other functions see: 
[https://ggplot2.tidyverse.org/reference/](https://ggplot2.tidyverse.org/reference/).

Besides, the cheetsheet of `ggplot2` can be downloaded as follow: 
[https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf](https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf). 
**HISS~**

# How to make simple plots

For showing how the `ggplot()` works, we will use the ggplot2 embedded data set `mpg` as a test 
data set. This data is about technical spec of cars. Here are some explanations for the 
attributions in this data set:    
1) manufacturer   
2) model > model name   
3) displ > engine displacement, in liters or size of engine   
4) year > year of manufacture   
5) cyl > number of cylinders    
6) trans > type of transmission   
7) drv > f = front-wheel drive, r = rear wheel drive, 4 = 4 wheel drive   
8) cty > city miles per gallon    
9) hwy > highway miles per gallon or efficiency   
10) fl > fuel type    
11) class > "type" of car   

```{r}
mpg
## If you want to look at the data in RStudio assign mpg in an object
mpg <- mpg
```

## Scatter plot

Here is an example for making a scatter plot.

```{r}
ggplot(data = mpg, aes(x = displ, ## value for x axis is displ
                       y = hwy,   ## value for y axis is hwy
                       color = class)) + ## color points by their class
  geom_point() ## the function for making a scatter plot
```

Shape the shape by the fuel type (fl).

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class, shape = fl)) + 
  geom_point() 
```

Add regression line to all the points.

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth(method='lm', formula=y~x) ## the function to add a regression line
```

As you can see up there, the grammar of `ggplot2` is simple. You can just input the data set, select 
`aesthetic` then add the plot that you want to make layer by layer with `+`.

##  Histogram

A histogram is an approximate representation of the distribution of numerical data. 

```{r}
## here we use a histogram to show the distribution of highway miles per gallon or efficiency
ggplot(data=mpg,aes(x=hwy)) + geom_histogram()
```

## Density plot

A density plot shows the distribution of numerical data.

```{r}
## here we use a density plot to show the distribution of the efficiency across fuel types
ggplot(data=mpg, aes(x=hwy, color = fl)) + geom_density()
```

## Barplot

The barplot can be used to compare the numbers among different groups.

```{r}
## here we use a barplot to show different manufactures
ggplot(data = mpg, aes(x = manufacturer)) + 
    geom_bar() + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) ## this command adjust the angle of text at the x-axis
```

## Boxplot

In descriptive statistics, a boxplot is a method for graphically demonstrating the locality, 
spread and skewness groups of numerical data through their quartiles. 

```{r}
## a boxplot for the hwy 
ggplot(mpg, aes(x = class, y  = hwy)) + 
    geom_boxplot()
```

## Violin plot

Another graphics useful to observe the distribution of data is a violin plot, a sort density plot 
that is rotated and placed on each side, to show the distribution shape of the data.

```{r}
## a violin plot for the hwy
ggplot(mpg, aes(x = class, y  = hwy)) + 
  geom_violin()  + 
  geom_boxplot(width = 0.1)
```
## Beeswarm plot
Yet another way to visualize a plot like this is the beeswarm method.

This can be helpful to show the number of observations per box/violin.

```{r}
library(ggbeeswarm)
ggplot(mpg, aes(x = class, y  = hwy)) + 
  geom_quasirandom()
```

# How to customize plots

Besides make easy plots with the defaults, we can also customize almost everything in the plot, 
e.g. the font size, color, text, labels and so on.

To explore the customize plot, let's first try to change the color in our violin plot. We can assign the plot in an object. e.g. `p1` for later usage.

```{r}
## assign the plot in an object for the later usage
p1 <- ggplot(mpg, aes(x = class, y  = hwy)) + 
    geom_violin(color = "blue")  + 
    geom_boxplot(width = 0.1, fill = "orange")
p1 ## calling the plot from the global environment
```

Now let's try to add a title to the plot and remove the label on x-axis.

```{r}
p1 + labs(title = "Box violin plot",
          x = NULL)
```

Besides manually change all the parameters and try to make the plot looks nicer, in the `ggplot2` 
package, they actually offer different themes which allow us to quickly change the style of the plot.

```{r}
p1 + theme_bw()
```

```{r}
p1 + theme_dark()
```
If you are interested in select a nice theme for your future plots, you can check out the following 
link [https://ggplot2.tidyverse.org/reference/ggtheme.html](https://ggplot2.tidyverse.org/reference/ggtheme.html).

We can also customize our theme further. There are many arguments with which you can change position, color and size the elements individually. [https://ggplot2.tidyverse.org/reference/theme.html](https://ggplot2.tidyverse.org/reference/theme.html)

```{r}
p1 + theme_dark() +
  theme(
  # get rid of panel grids
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank()
  )
```

It is also possible to filter your data directly in the ggplot function.

```{r}
p2 <- ggplot(data = mpg[mpg$displ <= 6, ], aes(x = displ, y = hwy, colour = year)) + 
  geom_point()
p2
```

There are multiple `scale_color_*` functions which can be used to define colors for the plot. It is important to choose a function that fits with data the color is mapped onto. Some functions work with continuous values while others work with discrete values.

```{r}
ggplot(data = mpg[mpg$displ <= 6, ], aes(x = displ, y = hwy, colour = year)) + 
  geom_point() +
  # since year is an integer column the values are treated as if on a continuous scale even though there are only two numbers
  scale_colour_gradient( low = "steelblue", high = "firebrick")

ggplot(data = mpg[mpg$displ <= 6, ], aes(x = displ, y = hwy, colour = fl)) + 
  geom_point() +
  scale_colour_manual(values = c("p" = "blue", "r" = "red", "e" = "green", "d" = "yellow", "c" = "magenta"))
```

## Making subplots

`ggplot2` offers a function `facet_wrap` that splits the data into multiple subplots.

```{r}
p3 <- ggplot(data = mpg, aes(x = displ, y = hwy)) + 
        geom_point() +
        geom_smooth(method='lm', formula=y~x)
p3

p3 + facet_wrap(~fl)
p3 + facet_wrap(~fl, scales = "free")
```

As you can see the two plots share the same x- and y-axis. We can set individual axis for each plot using `scales = "free"`. Take notice that the regression line is calculated and drawn for each subplot individually using only the data belonging to the subplot.

If you want to display the data from multiple columns in a faceted structure you can use `melt()` from the `reshape2` package to put all column names you want to show in one single column. With the arguments `id.vars` and `measure.vars` you can choose which columns to keep and which columns to melt.

```{r}
#| eval: false
install.packages("reshape2")
```
```{r}
library(reshape2)
```

```{r}
mpg_melt <- melt(mpg, id.vars = c("cty", "hwy", "displ"), measure.vars = c("model", "manufacturer"))

ggplot(mpg_melt, aes(x=hwy, y=value)) +
  geom_point() +
  facet_wrap(~variable, scales="free_y") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

# Exercises

1) Use a bar chart to find out the which `audi` `model` has the highest frequency in the `mpg` data 
set.    
2) Use a scatter plot and regression line to find out whether there is a relation between `cty` and `hwy`.
3) Use a density plot to show the distribution of `hwy` of each `fl`for each `class`. 
4) Please try to answer the question: which manufacture produce the low energy consumption (`cty`) `midsize` cars?

```{r include=FALSE}
## Answer to question 1
ggplot(mpg[mpg$manufacturer == "audi",], 
       aes(x = model)) + 
  geom_bar() + 
  geom_label(stat='count', aes(label=..count..), vjust=0.5) + 
  theme_bw()
```

```{r include=FALSE}
## Answer to question 2
ggplot(mpg, 
       aes(x = cty, y = hwy)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  theme_bw()
```
 
```{r include=FALSE}
## Answer to question 3
## Density plot like above with facet_wrap
ggplot(data=mpg, aes(x=hwy, color = fl)) +
  geom_density() +
  facet_wrap(~class)
```
 
```{r include=FALSE}
## Answer to question 4
ggplot(mpg[mpg$class == "midsize",],
       aes(x = manufacturer, y = cty)) + 
  geom_boxplot()  + 
  theme_bw()
```

# Bonus exercise: Real World Applications

```{r}
library(tidyverse)
library(kableExtra) # useful for making tables
```

1) Choose 2 data sets and make 3 plots for each of them.
2) Customize the plots

## Median age per district

[Download here!](https://opendata.wuerzburg.de/explore/dataset/stadtbezirke_medianalter/table/)

```{r}
age <- read.table("data/stadtbezirke_medianalter.csv", sep=";", header=T)
head(age) %>% kable() %>% kable_styling("striped") %>% scroll_box(width="100%")
```

## Trees in Würzburg

[Download here!](https://opendata.wuerzburg.de/explore/dataset/baumkataster_stadt_wuerzburg/table/)

```{r}
trees <- read.table("data/baumkataster_stadt_wuerzburg.csv", sep=";", header=T)
head(trees) %>% kable() %>% kable_styling("striped") %>% scroll_box(width="100%")
```

## Weather in Gerbrunn

[Download here!](https://opendata.wuerzburg.de/explore/dataset/wetterstation-gerbrunn/table/)

```{r}
weather <- read.table("data/wetterstation-gerbrunn.csv", sep=";", header=T)
head(weather) %>% kable() %>% kable_styling("striped") %>% scroll_box(width="100%")
```

## Pedestrian counts

[Download here!](https://opendata.wuerzburg.de/explore/dataset/passantenzaehlung_stundendaten/table/)

```{r}
pedestrians <- read.table("data/passantenzaehlung_stundendaten.csv", sep=";", header=T) %>% select(!c(GeoShape, Zonen))
head(pedestrians) %>% kable() %>% kable_styling("striped") %>% scroll_box(width="100%")
```

## Bicycle counts

[Download here!](https://opendata.wuerzburg.de/explore/dataset/fahrradzaehlung-tagesdaten-alle-zaehlstationen/table/)

```{r}
bikes <- read.table("data/fahrradzaehlung-tagesdaten-alle-zaehlstationen.csv", sep=";", header=T)
head(bikes) %>% kable() %>% kable_styling("striped") %>% scroll_box(width="100%")
```
