---
title: "Manipulating data with tidyverse"
date: last-modified
author: 
- name:  "Natascha Kleinermanns & Zarnack group"
title-block-banner: true
format: 
    html:
        theme: litera
        embed-resources: true
        fig-format: png
        self-contained: true
        code-fold: false
        toc: true
        number-sections: false
        anchor-sections: false
execute:
  include: true
  eval: true
  echo: true
  error: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 100
---

# What is tidyverse

Tidyverse is the name for a collection of R packages that are useful for data science. The most important ones are:

-   `tidyr` which helps you to clean up your data
-   `dplyr` which offers a grammar for data manipulation
-   `ggplot2` which helps you to create graphics that can be customized to your needs

You can find more information on: <https://tidyverse.org/packages/>.

First we need to install `tidyverse`, which will install all our included packages.

```{r, eval=FALSE}
install.packages("tidyverse")
```

Afterwards we need to load the packages we want

```{r, eval=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

In this case instead of loading all packages separately, we can simply call `library(tidyverse)`. This will load 9 core packages (`tidyr`, `dplyr`, `purrr`, `readr`, `ggplot2`, `stringr`, `lubridate`, `forcats`, `tibble`) of tidyverse.

```{r, message=TRUE}
library(tidyverse)
library(kableExtra)
```

To demonstrate an practice using tidyverse operations we will use publicly available data sets.

```{r, eval=FALSE}
install.packages("datasets")
```

We are using the airquality data set.

```{r}
data("airquality")
airquality %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

# Basic tidyverse functions

## filter

We can use the function `filter()` to filter rows.

```{r}
airquality_may <- filter(airquality, Month == 5)
airquality_may %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

It is possible to combine multiple conditions.

```{r}
windy_may_days <- filter(airquality, Month == 5 & Wind > 10)
windy_may_days %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

We can also filter for everything except our condition

```{r}
filter(airquality, Month != 5) %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

If we want to filter for multiple values within a column we can do this with logical operators

```{r}
filter(airquality, Month == 6 | Month == 8) %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

or we can use the `%in%` operator.

```{r}
filter(airquality, Month %in% c(6, 8)) %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

This operator can be negated as well.

```{r}
# it is important to place "!" right
# if you put directly in front of %in% or c() you will get an error
filter(airquality, !(Month %in% c(6, 8))) %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

### Exercise 1

Filter for data from september.

```{r, include=FALSE}
# Solution exercise 1 filter
filter(airquality, Month == 9) %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

## arrange

We can also arrange the rows in different orders. Choose a column an whether to arrange the rows in ascending or descending order.

```{r}
# arranging in ascending order is the default and does not need another function
arrange(airquality, Temp) %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
# for arranging in descending order use desc()
arrange(airquality, desc(Temp)) %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

### Exercise 2

Arrange the days in may in ascending order of wind strength.

```{r, include=FALSE}
arrange(airquality_may, Wind) %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

## select

With `select()` we can select columns.

```{r}
# this will keep only the columns Wind, Day and Temp
windy_may_days <- select(windy_may_days, c(Wind, Day, Temp))
windy_may_days %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

We can also select columns we don't want, by placing `-` before our vector. This will keep all columns but Wind.

```{r}
windy_may_days <- select(windy_may_days, -c(Wind))
windy_may_days %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

### Exercise 3

Select Ozone and Solar.R from airquality.

```{r, include=FALSE}
select(airquality, c(Ozone, Solar.R)) %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

## mutate

`dplyr` also allows us to make new columns with `mutate()`, either with mathematical calculations or with logical operators.

```{r}
## Converting Fahrenheit to Celsius
airquality <- mutate(airquality, Temp_C = (Temp-32)*5/9)
airquality %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
## Defining days with Wind >= 10 as days with strong wind
airquality <- mutate(airquality, strong_wind = case_when(
  Wind >=10 ~ T,
  T~F
))
airquality %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

### Exercise 4

Make another column high_temp for temperatures equal to or above 25°C.

```{r, include=FALSE}
mutate(airquality, high_temp = case_when(
  Temp_C >= 25 ~ T,
  T~F
)) %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

## rename and relocate

We can rename and reorder columns, with `rename()` and `reorder()`. The syntax for renaming is new_name = old_name.

```{r}
airquality <- rename(airquality, Temp_F = Temp)
airquality %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

When reordering use `.before` or `.after` to put your moving column directly before or after another column.

```{r}
airquality <- relocate(airquality, Temp_C, .after = Temp_F)
airquality %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

### Exercise 5

Move the column wind between Ozone and Solar.R

```{r, include=FALSE}
relocate(airquality, Wind, .before = Solar.R) %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

## summerize

The dplyr package offers a function called `summarize()`. It works a bit differently than `summary()` from base R. For example applying it on our data frame without further specification returns an empty data frame.

```{r}
base::summary(airquality)
dplyr::summarize(airquality)
```

When using summarize we choose a column to summarize and a function to summarize that column with.

```{r}
summarize(airquality, mean(Temp_C), mean(Temp_F), median(Temp_C), , max(Temp_C), sum(strong_wind))
```

If we want to do the same operation on multiple columns we can use `across()`.

```{r}
summarize(airquality, across(c(Temp_C, Temp_F), mean))
```

We can also give names to our summary columns.

```{r}
summarize(airquality, mean = mean(Temp_C), median = median(Temp_C))
```

In the case that our data may contain NA values we can use the `na.rm =` argument. If there are no NAs this argument does not change the results.

```{r}
summarize(airquality, across(c(Temp_C, Temp_F), mean, na.rm = TRUE))
```

### Exercise 6

Summarize the mean Solar.R and Ozone

```{r, include=FALSE}
summarize(airquality, across(c(Ozone, Solar.R), mean, na.rm = TRUE))
```

## The pipe operator

The tidyverse package offers a pipe operator `%>%`, that can be used to make a chain of functions. When using `%>%` the data is passed from one function to the next along the pipeline.

```{r}
airquality_may <- airquality_may %>%
  mutate(Temp_C = (Temp-32)*5/9) %>% 
  mutate(strong_wind = case_when(
    Wind >=10 ~ T,
    T~F
  )) %>%
  rename(Temp_F = Temp) %>%
  relocate(Temp_C, .after = Temp_F)

airquality_may %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

The piping operator is also helpful for adjusting our `summerize()` function.

```{r}
airquality %>% 
  ## we can use a column to split our data in different groups first
  group_by(Month) %>%
  ## and then specify which columns we want to summarize
  summarize(across(c(Temp_C, Temp_F), mean)) %>% 
  # this function stops any grouping
  # it is used to explicitly make sure our df goes back to an ungrouped state
  ungroup()
```

### Exercise 7

Make a pipeline to calculate the mean Wind and Solar.R in june.

```{r, include=FALSE}
airquality %>%
  filter(Month == 6) %>%
  summarize(mean(Wind), mean(Solar.R))
```

# The melt function

`reshape2` is another useful package next to tidyverse. It offers the `melt()` function, which will melt all columns together. This results in a data frame with one variable column with the former column names and one value column with the former column values. With the argument `id.vars =` we can exclude columns from being melted and with `measure.vars =` we can specifically collect columns to be melted. With `na.rm =` we can choose whether to keep rows where the value is NA or drop them.

```{r eval=FALSE}
install.packages("reshape2")
```
```{r}
library(reshape2)
```

```{r}
# if measure.vars is not specified all columns are melted, except the ones selected by id.vars
air_melt <- melt(airquality, id.vars = c("Month", "Day"), variable.name = "air_factor", value.name = "measured")
air_melt %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

### Exercise 8

Melt the Ozone, Solar.R and Wind columns in airquality_may.

```{r include=FALSE}
may_melt <- melt(airquality_may, measure.vars = c("Ozone", "Solar.R", "Wind"))
may_melt %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

# Excercises Tidyverse

1) What is the hottest day in each month?

```{r, include=FALSE}
airquality %>%
  group_by(Month) %>%
  summarize(max(Temp_F))
```

2) Add a column hot_and_windy

```{r, include=FALSE}
airquality <- airquality %>%
  mutate(hot_and_windy = case_when(
    strong_wind == TRUE & Temp_C >= 25 ~ TRUE,
    T~F
  ))
airquality %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

3) Sort days with strong wind after temperature in ascending order

```{r, include=FALSE}
airquality %>%
  filter(strong_wind) %>%
  arrange(Temp_C) %>% kable() %>% kable_styling("striped") %>% scroll_box(height = "400px")
```

# Bonus Excercise Tidyverse & ggplot2
1)  Plot the wind on each day in may (Month 5) and color wind intensities above 10 blue

```{r include=FALSE}
ggplot(airquality[airquality$Month==5,], aes(x=Day, y=Wind, fill = strong_wind))+
  geom_col() +
  scale_fill_manual(values =c("grey", "blue"))
  
```

3)  Use facet_wrap to plot the wind in each month

```{r include=FALSE}
ggplot(airquality, aes(x=Day, y=Wind, fill = strong_wind))+
  geom_col() +
  scale_fill_manual(values =c("grey", "blue")) +
  facet_wrap(~Month)
```

4)  Plot only ozone, solar.r wind and Temp_C eafor month 5 as facets of the same plot

```{r include=FALSE}
melt_df <- melt(airquality, id.vars = c("Month", "Day"), measure.vars = c("Ozone", "Solar.R", "Temp_C", "Wind"))
ggplot(melt_df[melt_df$Month == 5, ], aes(x = Day, y = value)) +
  geom_point() +
  geom_line() +
  facet_wrap(~variable, scales = "free_y")
```

5)  Make a plot with panels for each condition and month with `facet_grid()`.

```{r include=FALSE}
ggplot(melt_df, aes(x = Day, y = value)) +
  geom_point() +
  geom_line() +
  facet_grid(variable~Month, scales = "free")
```

6)  Plot the temperatures of each month and color by the temperature (warmest red, coldest blue)

```{r include=FALSE}
ggplot(airquality, aes(x=Day, y=Temp_C, fill = Temp_C))+
  geom_col()+
  facet_wrap(~Month, ncol = 5)+
  scale_fill_gradient(low="blue", high="red")
```

# Session info

```{r}
sessionInfo()
```