---
title: "Debugging in R: How to Read and Fix Error Messages"
date: last-modified
author: 
- name:  "Melina Klostermann"
title-block-banner: true
format: 
    html:
        theme: litera
        self-contained: true
        code-fold: false
        toc: true
        number-sections: false
        anchor-sections: false
execute:
  include: true
  eval: true
  echo: true
  error: true
editor: 
  markdown: 
    wrap: 100
---

# Learning goals

After this session you should be able to:

* read and interpret R error messages
* identify common causes of errors
* use `?function` to read documentation
* search for solutions online
* ask clear programming questions

---

# 1. Errors are Normal

Even experienced programmers constantly see errors.

Errors usually mean one of three things:

1. R **cannot find something**
2. R **received the wrong type of input**
3. R **the syntax is incorrect**

The key skill is **reading the error message carefully**.

---

# 2. Anatomy of an Error Message

Run the following code:

```{r}
x <- c(1,2,3)
mean(xyz)
```

Example error:

```
Error in mean(xyz) : object 'xyz' not found
```

How to read this:

| Part                     | Meaning                               |
| ------------------------ | ------------------------------------- |
| `Error in mean(xyz)`     | the function where the error happened |
| `object 'xyz' not found` | the actual problem                    |

**Interpretation:**
R cannot find an object named `xyz`.

**Solution:**
Use the same object name as you used before.
```{r}
x <- c(1,2,3)
mean(x)
```

---

**Exercise 1: Fix the Error**

```{r}
numbers <- c(3,7,10,2)
mean(number)
```

What is the problem? And what would be the correct code?

---

# 3. Typical Errors

---

## 3.1 Object Not Found

This is one of the **most common beginner errors**.

```{r}
y + 5
```

Why does this fail?

<details>
<summary>Solution</summary>

`y` has never been defined.

Define it first:

```r
y <- 10
y + 5
```

</details>

---

**Exercise 2: Fix the Object Not Found Error**

What is wrong with this code? Fix it so it runs without an error.

```{r}
my_values <- c(4, 8, 15, 16, 23, 42)
sum(my_value)
```

---

**Exercise 3: Define the Missing Object**

This code produces an error because something is missing. Add the necessary line to make it work.

```{r}
result <- total_count / 100
```

---

## 3.2 Syntax Errors

A syntax error means R cannot understand the **structure** of the code — something is missing or in the wrong place.

```{r}
x <- c(1, 2, 3
mean(x)
```

Typical message:

```
Error: unexpected symbol in "mean"
```

Common causes:

| Problem              | Example                    |
|----------------------|----------------------------|
| Missing `)`          | `c(1, 2, 3`                |
| Missing `]`          | `df[1, 2`                  |
| Missing comma        | `c(1 2 3)`                 |
| Unclosed quote       | `x <- "hello`              |
| Missing `+` in ggplot| Next line starts with `geom_`|

Tip: R Studio highlights unclosed brackets and quotes. Look for the red underlines or unexpected cursor position.

---

**Exercise 4: Fix the Syntax Errors**

Each code block below contains a syntax error. Find and fix it.

**Case 1:**

```{r}
my_vec <- c(10, 20, 30
sum(my_vec)
```

**Case 2:**

```{r}
df <- data.frame(
  gene = c("A", "B", "C")
  value = c(1, 2, 3)
)
```

**Case 3:**

```{r}
greeting <- "Hello
print(greeting)
```

---

## 3.3 Package or Library Not Found

Another very common "not found" error occurs when you try to use a function from a package that is not available.

```{r}
library(skimr)
```


This means either the package is **not installed**, or it is installed but **not loaded**.


**Installing vs. Loading**

| Action      | Command                    | When needed                        |
|-------------|----------------------------|------------------------------------|
| **Install** | `install.packages("pkg")`  | Once, to download the package      |
| **Load**    | `library(pkg)`             | Every R session where you use it   |

Installing puts the package on your computer. Loading makes it available in the current session.

A common mistake is to install a package but forget to load it:

```{r}
# Installed but not loaded — this will fail:
skim(data.frame(x = c(1, 2, 3), y = c(4, 5, 6)))
```

`skimr` is a useful package that gives a quick summary of a data frame.

Fix: always install & load the package first. 

```{r}
install.packages("skimr")
library(skimr)
skim(df)
```

Note: You only need to install a package once on your computer. But you need to load the library everytime you open Rstudio or a new session.

**Exercise 5: Why is it working now?**
```{r}
skim(df)
```

---

**CRAN vs. Bioconductor**

There are two main repositories for R packages:

**CRAN** (Comprehensive R Archive Network)

* The default repository for general-purpose R packages
* Install with `install.packages("packagename")`

**Bioconductor**

* A specialized repository for **bioinformatics** packages (e.g. DESeq2, edgeR, GenomicRanges)
* Has its own installer and release cycle tied to R versions
* Install with the `BiocManager` package:

```r
install.packages("BiocManager")
BiocManager::install("packagename")
```

---

**Example: `install.packages()` Fails for a Bioconductor Package**

Try running:

```{r}
install.packages("DESeq2")
```

You will likely see a warning like:

```
Warning: package 'DESeq2' is not available for this version of R
```

This happens because `DESeq2` is a **Bioconductor** package, not a CRAN package.

The correct way to install it:

```r
install.packages("BiocManager")
BiocManager::install("DESeq2")
```
---

**Exercise 6: CRAN or Bioconductor?**

For each package below, search online to find out whether it is available on **CRAN** or **Bioconductor**, then write the correct install command.

| Package        | Repository | Install command |
|----------------|------------|-----------------|
| `dplyr`        |            |                 |
| `edgeR`        |            |                 |
| `ggplot2`      |            |                 |
| `GenomicRanges`|            |                 |
| `stringr`      |            |                 |

Tip: search for `R package <packagename>` and check whether the result links to `cran.r-project.org` or `bioconductor.org`.

---

## 3.4 Function Not Found

```{r}
mutate(df, hello = "hello")
```



This error has exactly **two possible causes**:

1. There is a **typo** in the function name
2. The **package containing the function is not loaded**

These look identical from the error message alone, so you need to investigate.

**How to tell them apart:**

* Search the function name with `?functionname` or online — if the function exists in a package that you have not loaded, the help page will either not appear or will tell you which package it belongs to.
* Check your `library()` calls at the top of your script.

---

**Example A: Typo in the function name**

```{r}
x <- c(1, 3, 5, 7)
squrt(x)
```

The function `squrt` does not exist. The correct name is `sqrt`.

---

**Example B: Package not loaded**

```{r}
counts <- data.frame(gene = c("A", "B", "C"), value = c(10, 5, 8))
mutate(counts, log_value = log(value))
```

`mutate` exists — it is part of the `dplyr` package — but the package has not been loaded.

Fix:

```r
library(dplyr)
mutate(counts, log_value = log(value))
```

---

**Exercise 7: Typo or Missing Library?**

For each error below, decide: is it a **typo** or a **missing library**? Write down your reasoning and the fix.

**Case 1:**

```{r}
x <- c(2, 4, 6, 8)
Maen(x)
```

**Case 2:**

```{r}
df <- data.frame(a = c(1, 2, 3), b = c(4, 5, 6))
select(df, a)
```

**Case 3:**

```{r}
x <- c(10, 20, 30)
suM(x)
```

Tip: use `?functionname` to check whether a function exists. If the help page opens, the function is real — you may just need to load its package.

---

## 3.5 Reading Documentation

You can open help pages directly in R.

Example:

```{r}
?mean
```

Look for these sections:

| Section     | Meaning                |
| ----------- | ---------------------- |
| Description | what the function does |
| Usage       | how to call it         |
| Arguments   | explanation of inputs  |
| Examples    | runnable examples      |

---

**Exercise 8: Using Documentation**

Find out:

* What does `sd()` do?
* What input does it require?

Try:

```{r}
?sd
```

Then run:

```{r}
sd(c(1,4,6,10))
```

---

**Exercise 9: Explore Three Functions**

Use `?` to open the help page for each function below. For each one, write down:

* What does it do?
* What input does it require?
* Run the example code and describe what the output means.

**Function 1:** `round()`

```{r}
?round
round(3.14159, digits = 2)
```

**Function 2:** `nchar()`

```{r}
?nchar
nchar("Hello World")
```

**Function 3:** `seq()`

```{r}
?seq
seq(from = 1, to = 10, by = 2)
```

---

## 3.6 Wrong Function Inputs

Functions can fail not only because of typos or missing packages, but also because you give them **the wrong inputs**. There are two common ways this happens:

1. **Wrong column name** — you access a column that does not exist
2. **Wrong argument name** — you use an argument that the function does not recognise

---

### 3.6.1 Wrong Column Name

Data frame column names must match exactly — R is **case sensitive**.

```{r}
df <- data.frame(
  gene = c("A", "B", "C"),
  expression = c(5, 7, 3)
)

mean(df$Expression)
```

The column is called `expression`, not `Expression`. R returns `NULL` silently or an error depending on the context.

Tip: always check column names first:

```{r}
names(df)
```

---

### 3.6.2 Wrong Argument Name

Functions have specific argument names. Using the wrong name causes R to ignore the argument or throw an error.

```{r}
x <- c(1, 5, NA, 9)
mean(x, na_remove = TRUE)
```

Typical message:

```
Warning: In mean.default(x, na_remove = TRUE) :
  argument 'na_remove' matched by none
```

The correct argument is `na.rm`, not `na_remove`. Use `?mean` to check.

```r
mean(x, na.rm = TRUE)
```

---

**Exercise 10: Fix the Wrong Inputs**

Each example below has a wrong input. Identify the problem and fix it.

**Case 1:** Wrong column name

```{r}
counts_df <- data.frame(
  SampleID = c("S1", "S2", "S3"),
  ReadCount = c(1200, 850, 940)
)

mean(counts_df$readcount)
```

**Case 2:** Wrong argument name

```{r}
x <- c(3, 7, NA, 2, 5)
sum(x, remove.na = TRUE)
```

Hint: use `?sum` to find the correct argument name.

**Case 3:** Column does not exist at all

```{r}
genes <- data.frame(
  name = c("TP53", "BRCA1"),
  logFC = c(1.2, -0.8)
)

genes$pvalue
```

What does R return when you access a column that does not exist? Is this an error, a warning, or something else?

---

## 3.7 Wrong Input Type

Some functions require numbers, others text, others data frames, and so on. Passing the wrong type causes an error.

```{r}
sqrt("hello")
```


This means that `sqrt()` expected a **numeric** input, but received **text** instead.

**Useful helper functions to check what type you have**

```{r}
x <- "hello"
class(x)
is.numeric(x)
str(x)
```

| Function       | What it tells you                              |
|----------------|------------------------------------------------|
| `class(x)`     | the type of the object (e.g. `"character"`, `"numeric"`, `"data.frame"`) |
| `is.numeric(x)`| `TRUE` if numeric, `FALSE` otherwise           |
| `str(x)`       | a compact overview of structure and type       |

Common types in R:

| Type        | Example                     |
|-------------|-----------------------------|
| `numeric`   | `x <- 3.14`                 |
| `character` | `x <- "hello"`              |
| `logical`   | `x <- TRUE`                 |
| `data.frame`| `x <- data.frame(a = 1:3)`  |

---

**Exercise 11: What Type Is It?**

For each object below, predict what `class()` will return, then run the code to check.

```{r}
a <- 42
b <- "42"
c <- TRUE
d <- c(1, 2, 3)
e <- data.frame(x = 1:3)

class(a)
class(b)
class(c)
class(d)
class(e)
```

Now try using `a` and `b` with `sqrt()`. Which one throws an error and why?

```{r}
sqrt(a)
sqrt(b)
```

---

**Exercise 12: Fix the Type Error**

The code below produces a type error. Use `class()` to diagnose the problem, then fix it.

```{r}
counts <- c("10", "20", "30")
mean(counts)
```

Hint: look at what `class(counts)` returns. To convert text to numbers use `as.numeric()`.

---

## 3.8 File or Path Not Found

A very common error when reading data files is:

```
Error in file(file, "rt") : cannot open the connection
In addition: Warning message:
In file(file, "rt") : cannot open file 'data/counts.csv': No such file or directory
```

This means R cannot find the file at the path you provided.

**Common causes:**

| Cause | Example |
|---|---|
| Wrong file name | `counts.csv` instead of `Counts.csv` |
| Wrong folder | file is in `data/` but you wrote `input/` |
| Working directory is not where you think | R is looking in a different folder |
| Path separators wrong on Windows | `data\counts.csv` instead of `data/counts.csv` |

**Check your working directory:**

```{r}
getwd()
```

This shows where R is currently looking for files. Make sure your file path is relative to this location, or use an absolute path.

**Windows users: slash direction**

On Windows, file paths in File Explorer use backslashes:

```
C:\Users\student\data\counts.csv
```

R requires **forward slashes** or **double backslashes**:

```r
# These both work in R on Windows:
read.csv("C:/Users/student/data/counts.csv")
read.csv("C:\\Users\\student\\data\\counts.csv")

# This does NOT work:
read.csv("C:\Users\student\data\counts.csv")
```

**Tip:** In RStudio, use the **Files** pane to navigate to your file, then click *More > Set As Working Directory* to set the right working directory automatically.

**Exercise 13: Fix the File Path**

The code below throws a "cannot open file" error. What are three things you would check to fix it?

```{r}
counts <- read.csv("Data/gene_counts.CSV")
```

---

# 4. Error Solving Strategies

---

## 4.1 Strategy for Debugging

When you see an error:

1. **Read the error message carefully**
2. Identify **which function caused it**
3. Check:

   * variable names
   * column names
   * argument names
4. Look at documentation

---

## 4.2 Searching for Errors Online

Most programming problems have already been solved.

Good places to search:

* Stack Overflow
* Bioconductor Support
* RStudio Community
* package documentation

Good search query example:

```
R error "object not found" dplyr mutate
```

Include:

* the **exact error message**
* the **function name**
* the **package name**

---

## 4.3 Asking Good Questions

When asking for help include:

1. the **exact error message**
2. the **code that caused the error**
3. a **minimal example**
4. what you already tried

Example:

Bad question:

> My R code does not work.

Better question:

> I get the error `object 'gene' not found` when running:
>
> ```r
> mean(df$gene)
> ```
>
> My data frame is:
>
> ```r
> df <- data.frame(expression = c(3,4,5))
> ```

---

## 4.4 Using ChatGPT for Debugging

When asking an AI assistant include:

* the full error message
* the code
* expected output

Example prompt:

```
I get this error in R:

Error in mean(number) : object 'number' not found

My code is:

numbers <- c(3,7,10)
mean(number)

What causes this error?
```

---

## 4.5 Final Exercise: Debug the Script

This script contains multiple errors.

```{r}
data <- data.frame(
  gene = c("A","B","C"),
  counts = c(10,5,7)
)

mean(data$count)

sd(data$counts

sqrt("25")
```

Tasks:

1. run the code
2. read the errors
3. fix all problems

<details>
<summary>Solution</summary>

```r
data <- data.frame(
  gene = c("A","B","C"),
  counts = c(10,5,7)
)

mean(data$counts)

sd(data$counts)

sqrt(25)
```

</details>

---

## 4.6 Final Exercise: Wrong Package and Wrong Column

The script below has two problems: a package issue and a column name issue. Fix both.

```{r}
library(dplyr)

patients <- data.frame(
  PatientID = c("P1", "P2", "P3"),
  Age = c(34, 45, 29),
  Score = c(8.2, 6.5, 9.1)
)

filter(patients, age > 30)
```

Tasks:

1. Run the code and read the error
2. Identify both problems
3. Fix the script so it runs correctly

<details>
<summary>Solution</summary>

```r
# dplyr is already loaded so that is fine.
# The column is called "Age" not "age" — column names are case sensitive.

library(dplyr)

patients <- data.frame(
  PatientID = c("P1", "P2", "P3"),
  Age = c(34, 45, 29),
  Score = c(8.2, 6.5, 9.1)
)

filter(patients, Age > 30)
```

</details>

---

## 4.7 Final Exercise: Find the Error Type

For each error message below, identify: what **type** of error is it, and what is the most likely fix?

**Error 1:**
```
Error: object 'gene_counts' not found
```

**Error 2:**
```
Error in library(DESeq2) : there is no package called 'DESeq2'
```

**Error 3:**
```
Error in sqrt(x) : non-numeric argument to mathematical function
```

**Error 4:**
```
Error: unexpected symbol in "mean(x"
```

**Error 5:**
```
Error: could not find function "ggplot"
```

For each one, write:

- The error type (from section 3)
- A likely cause
- The fix you would try first

---

# Take-home Message

Programming is not about avoiding errors.

It is about **learning how to interpret them quickly**.
