---
title: "Debugging in R: Solutions"
date: last-modified
author: 
- name:  "Melina Klostermann"
title-block-banner: true
format: 
    html:
        theme: litera
        self-contained: true
        code-fold: false
        toc: true
        number-sections: false
        anchor-sections: false
execute:
  include: true
  eval: true
  echo: true
  error: true
editor: 
  markdown: 
    wrap: 100
---

This file contains solutions for all exercises from `03_Debugging_in_R.qmd`.

---

# Exercise 1: Fix the Error

**Problem:** The object is named `numbers` but `mean(number)` uses `number` — a typo (missing `s`).

```{r}
numbers <- c(3, 7, 10, 2)
mean(numbers)
```

---

# Exercise 2: Fix the Object Not Found Error

**Problem:** The object is named `my_values` but `sum(my_value)` is missing the final `s`.

```{r}
my_values <- c(4, 8, 15, 16, 23, 42)
sum(my_values)
```

---

# Exercise 3: Define the Missing Object

**Problem:** `total_count` has never been defined. A value must be assigned first.

```{r}
total_count <- 500   # define the missing object with any appropriate value
result <- total_count / 100
result
```

---

# Exercise 4: Fix the Syntax Errors

**Case 1:** Missing closing parenthesis `)` in `c(10, 20, 30`.

```{r}
my_vec <- c(10, 20, 30)
sum(my_vec)
```

**Case 2:** Missing comma `,` after the first element in `data.frame()`.

```{r}
df <- data.frame(
  gene = c("A", "B", "C"),
  value = c(1, 2, 3)
)
```

**Case 3:** Unclosed string — the opening quote `"` for `"Hello` is never closed.

```{r}
greeting <- "Hello"
print(greeting)
```

---

# Exercise 5: Why Is It Working Now?

`skim(df)` works here because `library(skimr)` was called earlier in the document, which loaded the `skimr` package into the current R session. Once a package is loaded with `library()`, all its functions — including `skim()` — are available for the rest of the session.

```{r}
library(skimr)
df <- data.frame(x = c(1, 2, 3), y = c(4, 5, 6))
skim(df)
```

---

# Exercise 6: CRAN or Bioconductor?

| Package         | Repository    | Install command                                 |
|-----------------|---------------|-------------------------------------------------|
| `dplyr`         | CRAN          | `install.packages("dplyr")`                     |
| `edgeR`         | Bioconductor  | `BiocManager::install("edgeR")`                 |
| `ggplot2`       | CRAN          | `install.packages("ggplot2")`                   |
| `GenomicRanges` | Bioconductor  | `BiocManager::install("GenomicRanges")`         |
| `stringr`       | CRAN          | `install.packages("stringr")`                   |

Reminder: to use `BiocManager::install()`, first install BiocManager from CRAN if you have not already:

```r
install.packages("BiocManager")
```

---

# Exercise 7: Typo or Missing Library?

**Case 1:** `Maen(x)` — **Typo**. The function `Maen` does not exist. The correct name is `mean` (lowercase).

```{r}
x <- c(2, 4, 6, 8)
mean(x)
```

**Case 2:** `select(df, a)` — **Missing library**. `select()` is a real function from the `dplyr` package, but the package has not been loaded.

```{r}
library(dplyr)
df <- data.frame(a = c(1, 2, 3), b = c(4, 5, 6))
select(df, a)
```

**Case 3:** `suM(x)` — **Typo**. R is case-sensitive; `suM` does not exist. The correct name is `sum` (all lowercase).

```{r}
x <- c(10, 20, 30)
sum(x)
```

---

# Exercise 8: Using Documentation

`sd()` computes the **standard deviation** of a numeric vector. It requires a numeric vector as input (`x`). The optional argument `na.rm = TRUE` removes missing values before calculation.

```{r}
?sd
sd(c(1, 4, 6, 10))
```

The result is approximately `3.83` — telling us how spread out the values are around their mean.

---

# Exercise 9: Explore Three Functions

**Function 1: `round()`**

* **What it does:** rounds numbers to a specified number of decimal places.
* **Input required:** a numeric value or vector (`x`), and optionally the number of decimal places (`digits`, default `0`).
* **Output:** `round(3.14159, digits = 2)` returns `3.14` — the number rounded to 2 decimal places.

```{r}
?round
round(3.14159, digits = 2)
```

---

**Function 2: `nchar()`**

* **What it does:** counts the number of characters in each element of a character vector.
* **Input required:** a character vector (`x`).
* **Output:** `nchar("Hello World")` returns `11` — the number of characters (including the space) in the string.

```{r}
?nchar
nchar("Hello World")
```

---

**Function 3: `seq()`**

* **What it does:** generates a regular sequence of numbers.
* **Input required:** `from` (start value), `to` (end value), and either `by` (step size) or `length.out` (total number of values).
* **Output:** `seq(from = 1, to = 10, by = 2)` returns `1 3 5 7 9` — a sequence starting at 1, ending at or before 10, with steps of 2.

```{r}
?seq
seq(from = 1, to = 10, by = 2)
```

---

# Exercise 10: Fix the Wrong Inputs

**Case 1:** Wrong column name — `readcount` should be `ReadCount` (R is case-sensitive).

```{r}
counts_df <- data.frame(
  SampleID = c("S1", "S2", "S3"),
  ReadCount = c(1200, 850, 940)
)

mean(counts_df$ReadCount)
```

**Case 2:** Wrong argument name — `remove.na` is not an argument of `sum()`. The correct argument is `na.rm`.

```{r}
x <- c(3, 7, NA, 2, 5)
sum(x, na.rm = TRUE)
```

**Case 3:** `genes$pvalue` — the column `pvalue` does not exist in the data frame. R returns `NULL` silently (no error, no warning). This is a common source of confusion because the code appears to "work" but silently produces missing data.

```{r}
genes <- data.frame(
  name = c("TP53", "BRCA1"),
  logFC = c(1.2, -0.8)
)

genes$pvalue   # returns NULL — the column does not exist
names(genes)   # check which columns actually exist
```

---

# Exercise 11: What Type Is It?

Predicted and observed types:

| Object | Prediction    | `class()` result |
|--------|---------------|-----------------|
| `a`    | `"numeric"`   | `"numeric"`     |
| `b`    | `"character"` | `"character"`   |
| `c`    | `"logical"`   | `"logical"`     |
| `d`    | `"numeric"`   | `"numeric"`     |
| `e`    | `"data.frame"`| `"data.frame"`  |

```{r}
a <- 42
b <- "42"
c <- TRUE
d <- c(1, 2, 3)
e <- data.frame(x = 1:3)

class(a)
class(b)
class(c)
class(d)
class(e)
```

`sqrt(a)` works because `a` is numeric. `sqrt(b)` throws an error because `b` is a character string — `sqrt()` cannot take non-numeric input.

```{r}
sqrt(a)
sqrt(b)
```

---

# Exercise 12: Fix the Type Error

**Problem:** `counts` is a character vector (each element is in quotes). `mean()` requires numeric input.

```{r}
counts <- c("10", "20", "30")
class(counts)   # "character" — that is the problem
```

**Fix:** convert to numeric with `as.numeric()` first.

```{r}
counts <- c("10", "20", "30")
mean(as.numeric(counts))
```

---

# Exercise 13: Fix the File Path

Three things to check and fix in `read.csv("Data/gene_counts.CSV")`:

1. **Capitalisation of the folder name:** the folder may be called `data` (lowercase), not `Data`. File paths are case-sensitive on macOS and Linux.

2. **Capitalisation of the file extension:** the file extension may be `.csv` (lowercase), not `.CSV`. Again, case-sensitive on macOS/Linux.

3. **Working directory:** use `getwd()` to confirm that R is looking in the folder you expect. If not, adjust the path or change the working directory with `setwd()` or via the RStudio Files pane.

```r
getwd()   # check where R is looking

# Corrected version (assuming the folder and file use the correct capitalisation):
counts <- read.csv("data/gene_counts.csv")
```

---

# Exercise 4.5: Debug the Script

The original script has three errors:

1. `mean(data$count)` — the column is called `counts`, not `count`.
2. `sd(data$counts` — missing closing parenthesis `)`.
3. `sqrt("25")` — `"25"` is a character string; `sqrt()` requires a numeric value.

**Fixed script:**

```{r}
data <- data.frame(
  gene = c("A", "B", "C"),
  counts = c(10, 5, 7)
)

mean(data$counts)   # fix 1: correct column name

sd(data$counts)     # fix 2: added closing parenthesis

sqrt(25)            # fix 3: removed quotes — 25 is now numeric
```

---

# Exercise 4.6: Wrong Package and Wrong Column

**Problem 1:** The column is called `Age` (capital A), not `age`. Column names are case-sensitive in R.  
**Problem 2:** `dplyr` is already loaded via `library(dplyr)`, so that is not the issue here.

**Fixed script:**

```{r}
library(dplyr)

patients <- data.frame(
  PatientID = c("P1", "P2", "P3"),
  Age = c(34, 45, 29),
  Score = c(8.2, 6.5, 9.1)
)

filter(patients, Age > 30)   # fix: use "Age" with capital A
```

---

# Exercise 4.7: Find the Error Type

**Error 1:**
```
Error: object 'gene_counts' not found
```
- **Type:** Object not found (Section 3.1)
- **Likely cause:** The variable `gene_counts` was never assigned, or it was assigned with a different name.
- **Fix:** Define the object before using it: `gene_counts <- ...`

---

**Error 2:**
```
Error in library(DESeq2) : there is no package called 'DESeq2'
```
- **Type:** Package not found (Section 3.3)
- **Likely cause:** `DESeq2` is a Bioconductor package and has not been installed.
- **Fix:** Install via Bioconductor: `BiocManager::install("DESeq2")`, then `library(DESeq2)`.

---

**Error 3:**
```
Error in sqrt(x) : non-numeric argument to mathematical function
```
- **Type:** Wrong input type (Section 3.7)
- **Likely cause:** `x` is a character or logical vector instead of numeric.
- **Fix:** Check with `class(x)`; convert if needed: `sqrt(as.numeric(x))`.

---

**Error 4:**
```
Error: unexpected symbol in "mean(x"
```
- **Type:** Syntax error (Section 3.2)
- **Likely cause:** A closing parenthesis `)` is missing, or there is an unexpected character nearby.
- **Fix:** Complete the expression: `mean(x)`.

---

**Error 5:**
```
Error: could not find function "ggplot"
```
- **Type:** Function not found / missing library (Section 3.4)
- **Likely cause:** The `ggplot2` package has not been loaded in the current session.
- **Fix:** Add `library(ggplot2)` before calling `ggplot()`.
