---
title: "How to create a Quarto document?"
date: last-modified
author: 
    - name:  "Sri Sankaran"
title-block-banner: true
format: 
    html:
        theme: litera
        self-contained: true
        code-fold: false
        toc: true
        number-sections: false
        anchor-sections: false
execute:
  include: true
  eval: false
  echo: true
  error: false
editor: 
  markdown: 
    wrap: 100
---

# Quarto

Quarto is an open-source scientific and technical publishing system. Quarto files can be rendered
into HTML, PDF or word files as reports.

# Text formatting

-   *italics*, **bold**, ***bold italics***.
-   Programming commands presented exactly as written: `verbatim code`
-   Headings can be created using `#`.

# Execute codes

Note: The symbol `#` is used to add comments within a code chunk. Comments are not executed.

## Addition

```{r}
# Adding two numbers using R
1+1
```

## Subtraction

```{r}
# Subtracting two numbers using R
10-5
```

# Good Practices

Coding is not only about obtaining results, but also about producing results that are reproducible,
transparent, and well documented.

-   A Quarto document should always run from top to bottom without manual intervention.
-   Use meaningful section headings. Clear headings make your document easier to navigate and
    understand.
-   Avoid mixing unrelated tasks within the same code chunk.
-   One of the major strengths of R is the availability of numerous open-source packages for
    statistical and data analysis. These packages must be loaded before use. A good practice is to
    **create a dedicated setup chunk at the beginning of your document** where all required packages
    are loaded.
-   Packages are **installed once** using `install.packages()`, but must be **loaded in every new
    session** using `library()` function.
-   The same principles applies to parameters that will be reused throughout the analysis. For
    example, define a variable for pvalue cutoff like `pval -> 0.05` or predefined colors for
    conditions in plots or commonly used file paths. use during plotting.
-   Keep code clean and readable.
-   Add short comments where necessary to explain non-obvious steps.
-   Save and render your files frequently. This helps detect errors early and ensures that the
    analysis runs from start to finish without problems.
-   Use informative object names. (e.g., `normalized_read_counts_df` instead of `df1`)
-   When performing transformations or filtering, create new objects instead of modifying the
    original dataset. This preserves the original data and improves reproducibility.
-   Always use `sessionInfo()` at the end of your document to record the R versions and the versions
    of all loaded packages.

## Set up your document

At the beggining of the document, create a **setup code chunk** where you load, for example, all the
packages that will be used in the analysis.

```{r}
library(dplyr)
library(ggplot2)
```

## Name code chunks for easier navigation

Assign descriptive labels to code chunk to remember the purpose of each chunk for. For example, the
code chunk below is labeled colors because it defines colors that will later be used in plots. You
can write chunk options like code chunks labeliing using a sepcial comment symbol `#|`

```{r}
#| label: defining-colors

wt <- "grey"
mutant <- "purple"

```

## Customize your code chunks

When you render your document, you may want to control how code and its output appear in the final
report. This can be done by adding chunk options, which allow you to customize different components
of a code chunk.

For example, you may want to suppress messages or warnings that are not relevant for the reader.
This can be achieved using options such as message = FALSE or warning = FALSE, which prevent these
messages from appearing in the rendered document.

```{r}
#| label: loading_libraries
#| message: FALSE
library(dplyr)
library(ggplot2)
```

For more information about Quarto, visit: <https://quarto.org/docs/guide/>

# Session information

```{r}
sessionInfo()
```
